---
title: "Data Clean"
author: "Chandler Grote"
date: "2024-11-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
library(broom)
library(GGally)
library(tidyr)
library(tidymodels) 
library(kknn)
library(tidyverse)
library(ggh4x)
```

Each code block in this file should only need to run once. These are large csv's that are being loaded and this file is intended to avoid loading them every time during the rest of the project. When running a code chunk, run the entire chunk and wait for it to finish. Print statements have been added to the loops to track progress.

--NOTE: These are broken out for now but should be combined into one loop before final submission for run time and simplicity. Leave broken out for now as these chunks may need to be run separately if tweaks are made to the cleaning process before final submission.


## Clean each tracking week to only presnap events
```{r}

#Use a for loop to filter all 9 weeks of tracking data.
for (i in c(1:9)){
  #Add a print statement to track progress
  print(paste("Filtering Tracking Week", i))
  
  file_name <- paste0("data/tracking_week_", i, ".csv")
  
  #Create a dummy dataframe containing the current weeks tracking data
  temp <- read.csv(file_name)
  
  #We are only interested in *meaningful* pre-snap movement so remove all
  #AFTER_SNAP data and remove entries with minimal distance traveled.
  temp <- temp %>%   filter(frameType!="AFTER_SNAP", dis>0.1)
  
  #Write the filtered data to a new file
  write.csv(temp, paste0("data/presnap_events_week_", i, ".csv"), row.names = F)
}
```

## Find the frame where the line was set for each play
```{r}
print("Initialize master line set df with week 1")

file_name <- paste0("data/tracking_week_1.csv")

#Load in the first week of tracking data
temp <- read.csv(file_name)

#Create the master line set data frame
line_set_all <- temp %>%
  
  #Filter out the balls tracking data by removing NA values from the nflId
  #column. Filter by the line_set event so we can identify the appropriate frame
  filter(!is.na(nflId), event == "line_set") %>%
  
  #Group the data by game and play
  group_by(gameId, playId) %>%
  
  #Create a new variable that store the frameId of the line_set event
  reframe(line_set_frame = frameId) %>%
  group_by(gameId, playId) %>%
  
  #Remove the duplicate rows so we are left with a single line_set_frame value
  #for each play
  filter(!duplicated(line_set_frame)) %>%
  ungroup()

#Use a for loop to do the same process for the remaining weeks of tracking data
#and add them to the master line set data frame.
for (i in c(2:9)){
  #Add a print statement to track progress
  print(paste("Adding week", i, "to master line set df"))
  
  file_name <- paste0("data/tracking_week_", i, ".csv")
  
  temp <- read.csv(file_name)
  
  line_set_temp <- temp %>%
    filter(!is.na(nflId), event == "line_set") %>%
    group_by(gameId, playId) %>%
    reframe(line_set_frame = frameId) %>%
    group_by(gameId, playId) %>%
    filter(!duplicated(line_set_frame)) %>%
    ungroup()
  
  line_set_all <- line_set_all %>% rbind(line_set_temp)
}

#Write the master data frame to a new file for later use.
write.csv(line_set_all, paste0("data/tracking_line_set_frames.csv"), row.names = F)
```
