---
title: "Data Clean"
author: "Chandler Grote"
date: "2024-11-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
library(broom)
library(GGally)
library(tidyr)
library(tidymodels) 
library(kknn)
library(tidyverse)
library(ggh4x)
```

Each code block in this file should only need to run once. These are large csv's that are being loaded and this file is intended to avoid loading them every time during the rest of the project. When running a code chunk, run the entire chunk and wait for it to finish.

--NOTE: This file is intended to be ran top down, one chunk at a time. Previous chunks output is needed further down so run in order as they appear.

## Combining functions and cleaning methods into one process from other files.
Read the data files
```{r file-reading}
plays <- read.csv("data/plays.csv")

games <- read.csv("data/games.csv")

player_play <- read.csv("data/player_play.csv")

players <- read.csv("data/players.csv")
```


This output will be the final data frame used for modeling.

```{r plays-data-frame}
## Cleaning Plays data frame

# first get general df calculations and filtering.
plays_clean <- plays %>% filter(playNullifiedByPenalty == "N", # remove penalty plays
                                qbKneel == 0, # remove qb kneel plays
                                (dropbackType != "UNKNOWN" | is.na(dropbackType))) %>% # remove unknown dropbacks
  arrange(gameId, playId) %>% # sort by game and by play so each game is in order
  #first case when determines run pass on the play
  mutate(run_pass = case_when(pff_runPassOption == 1 ~ "run",
                              is.na(dropbackType) ~ "run",
                              str_detect(dropbackType, "RUN|SNEAK") ~ "run",
                              T ~ "pass"),
         # function to find how many total yards to opposing endzone
         yards_to_endzone = ifelse(possessionTeam == yardlineSide, abs(yardlineNumber - 100), yardlineNumber),
         # redzone indication
         redzone = ifelse(yards_to_endzone <= 20, T, F),
         # indication for which half of the game the play is in
         half = ifelse(quarter <= 2, 1, 2),
         # how many total seconds in the half
         # this is useful as the ball changes possession b/t halves but not quarters
         # --NOTE: do not break up following summation. 
         # It runs off the page but R doesn't like doing arithmetic on multiple lines
         sec_in_half = ((as.numeric(str_extract(gameClock, "[:digit:]+(?=:)")) * 60) + as.numeric(str_extract(gameClock, "(?<=:)[:digit:]+"))) + ifelse(quarter %in% c(1, 3), 900, 0))

# indication variables for a new half, game and possession
plays_clean$new_half <- plays_clean$half != c(0, plays_clean$half[-nrow(plays_clean)])
plays_clean$new_game <- plays_clean$gameId != c(0, plays_clean$gameId[-nrow(plays_clean)])
plays_clean$new_possession <- (plays_clean$possessionTeam != c(0, plays_clean$possessionTeam[-nrow(plays_clean)])) | plays_clean$new_half

# -- calculate the drive of the team. needed for EPA lag per drive.

# set first row of variables
teams <- c(plays_clean$possessionTeam[1], plays_clean$defensiveTeam[1]) # team names in game
drive <- c(1, 0) # which drive each team is on initially
names(drive) <- teams # name the drives vector for easy calling
plays_clean$drive <- c(1, rep(0, nrow(plays_clean) - 1)) # initial new column for drives

# loop through rows 2 to end of df
for(i in c(2:nrow(plays_clean))){
  # check for a new game
  if(plays_clean$new_game[i]){
    # update the following vectors with new game info
    teams <- c(plays_clean$possessionTeam[i], plays_clean$defensiveTeam[i])
    drive <- c(1, 0)
    names(drive) <- teams
  }
  # check for a new possession - includes a new half
  else if(plays_clean$new_possession[i]){
    # increment the drive of the current possession team
    drive[plays_clean$possessionTeam[i]] <- drive[plays_clean$possessionTeam[i]] + 1
  }
  # assign the drive number for the current possession team to the column in the df
  plays_clean$drive[i] <- drive[plays_clean$possessionTeam[i]]
}

# now that we have drives we can do EPA lag
plays_clean <- plays_clean %>%
  group_by(gameId, possessionTeam, drive) %>% # grouping
  mutate(epa_lag = c(0, expectedPointsAdded[-n()])) %>% # get the lag per drive starting at 0
  ungroup() %>% # ungroup
  arrange(gameId, playId) # arrange for easy viewing

# home, away, and week indications for each game
home_away <- games %>% select(gameId, week, homeTeamAbbr, visitorTeamAbbr)

# attributing win percentage for current possession team
plays_clean <- plays_clean %>% 
  # join the home_away df
  left_join(home_away, by = "gameId") %>%
  # attribute the correct pre-snap win percentage column to the team with the ball
  mutate(curr_win_percentage = ifelse(possessionTeam == homeTeamAbbr, preSnapHomeTeamWinProbability, preSnapVisitorTeamWinProbability))
```

```{r player-plays-data-frame}
## now cleaning player plays data frame

play_moves <- player_play %>% 
  # motionSinceLineset often contains the shiftsSinceLineset column
  # player_motion is defined when they are different to isolate motions from shifts
  mutate(player_motion = ifelse(shiftSinceLineset == motionSinceLineset, 0, 1)) %>% 
  group_by(gameId, playId) %>% # grouping by game then player
  left_join(players, by = "nflId") %>% # join players df to get positions
  summarise(
    # formation change happens when at least one shift occurs
    form_change = ifelse(sum(shiftSinceLineset, na.rm = TRUE) > 0, 1, 0),
    # Note: removed player_move line that was incomplete
    # player motion indicated when at least one of those players motions
    rb_motion = sum(position == "RB" & 
                      player_motion, na.rm = TRUE) > 0,
    te_motion = sum(position == "TE" & 
                      player_motion, na.rm = TRUE) > 0,
    wr_motion = sum(position == "WR" & 
                      player_motion, na.rm = TRUE) > 0,
    # same as player motion but only at the snap of the ball
    rb_atsnap = sum(position == "RB" & 
                      inMotionAtBallSnap, na.rm = TRUE) > 0,
    te_atsnap = sum(position == "TE" & 
                      inMotionAtBallSnap, na.rm = TRUE) > 0,
    wr_atsnap = sum(position == "WR" & 
                      inMotionAtBallSnap, na.rm = TRUE) > 0,
  ) %>%
  ungroup()

```

Merge cleaned data frames into final.
```{r final-df-merging}
# choose columns for modeling
final_df_columns <- c("gameId", "playId", "possessionTeam", "week", "run_pass", 
                      "down", "yardsToGo", "yards_to_endzone", "redzone",
                      "sec_in_half", "offenseFormation", "receiverAlignment", 
                      "epa_lag", "curr_win_percentage", "form_change",
                      "wr_motion", "te_motion","rb_motion",
                      "wr_atsnap", "te_atsnap", "rb_atsnap")

# join plays_clean and play_moves and select columns
final_df <- plays_clean %>%
  left_join(play_moves, by = c("gameId", "playId")) %>%
  select(all_of(final_df_columns)) %>%
  # change run_pass to a factor for classification modeling
  mutate(run_pass = as.factor(run_pass))

clean_column_names <- c("game_id", "play_id", "possession_team", "week", "run_pass", 
                      "down", "yards_to_go", "yards_to_endzone", "redzone",
                      "sec_in_half", "offense_formation", "receiver_alignment", 
                      "epa_lag", "curr_win_percentage", "form_change",
                      "wr_motion", "te_motion","rb_motion",
                      "wr_atsnap", "te_atsnap", "rb_atsnap")

colnames(final_df) <- clean_column_names


write.csv(final_df, "data/modeling_data_frame.csv", row.names = F)
```




