---
title: "NFL EDA - J"
author: "Justin Eldridge"
date: "2024-11-13"
output: pdf_document
---

# EDA for Justin to work on to avoid repo desync

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(broom)
library(GGally)
library(tidyr)
library(tidymodels) 
library(kknn)
library(tidyverse)
library(ggh4x)
library(dplyr)
```

#Try to select a model to predict pass vs. run

```{r}

#load in the tracking data for week 1
track1_df <- read.csv("data/presnap_events_week_1.csv")

#Explor the variables available to us
colnames(track1_df)

#Load in the plays and games data to explore the variables
plays_df <- read.csv("data/plays.csv")

#View the variables avaialble to get a sense if what to explore
colnames(plays_df)
```

## Brainstorm on variables to include

playid- to see what the previous play was to do a lag (i.e. if the previous run failed they are more liekly to do a pass)

time_remaining

yards remaining

score

maybe indicate if there was pre-snap movement 
  can create magnitude vectors and take a difference for each one
  use x coordinates
  
  orientation 360 where they ar facing 
  
distance direction and orientation for pre-snap movement 

If over a certain number indicate that pre-snap movement took place

want the sum of the distances traveled in each event before the snap

create our own distance column because they could still be moving even if their was no event. So when we removed the na values we removed some of the distances traveled before the snap

possibly filter out play with distance <0.1

Need to change the data cleaning file

to do lag on play we might have to sort by game ID, play id

red_zone 

possibly come back and make a function to calculate the teams record 

should exclude plays with a penalty would introduce too much error

exclude qbkneels, 





Indicator for significant pre-snap movement
time remaining
yards remaining
score
red-zone
play_id
game_id
play-concept lag
See if the teams current record is available
prensnap home and visitor score from the plays data
designed sneaks included don't include forced sneaks might be captured in run
runpassoption

either have to exclude rpo (don't happen often enough), or can include in prediction model as either run pass or rpo
  typcially fairly rare
  
  svm wouldnt be good for 3 classifiers, best for binary classifier
  
  or consider RPO vs run run 1st pass 2nd if the run isnt there then throw i
  so we would consider this a run play 
  Consider and rpo to be a run play
  if we do this just call them runs and then we could still use svm for the binary run pass classifier
  
  every rpo has a primary run concept pff thingy so we could use that column to classifiy 
  
  Break the data down into individual drives otherwise an epa lag will consider the other teams score in the new drive. 
  


rpo the qb chooses as the play unfolds whether hes going to pass or run



pass_length to try and get and indicator varialbe
could use rush_location_type

runconcept primary sometiems show as having one even if its a pass
  bc of designed rollout in dropback type have a run concept associated for some reason
  
  can use droback type as better indicator of pass
    -traditional scrabmle for passes, desgined rollouts-passes, 
    elminiate unknown dropback types
    
calculate yard to endzone because data fails to account for the 10 yards in the endzone(based on yard number)
-dont use yardstogo



2nd draft of variables:

  time remaining
  score
  yards remaining
  epa lag (break down into drives so lag works properly)
  run/pass indicator (include runpass options as runs so we can do svm,
  designed qb sneaks as runs scrambles as passes)
  play-concept lag
  red_zone
  game_id
  play_id
  indicator var for significant pre-snap movement
  pre-snap home and visitor score
  
  --they have var with game clock which resets every quarter
  might be able to make a function that takes game clock and gives us the amount of seconds remaining until end of the half
  
  
  
  to do:
  
  some eda
  
  how man runs vs. passes
  how often certain events occur look for rare events
  how many rpos end up as passes
  
  
  do svm for offense
  
  might do random forrest for defense
  two binary classifiers create 4 possile situtations
  
  
  data cleaning:
  
  -- make a function to convert the game clock into seconds remaining in the half.
  --play data, time on clock of play in minutes seconds.
  
  work on cleaning for the presnap events-
  use game id and play id to join the data, to attach cleaned data to the plays data
  
  neeed out of presnap, which player is moving on the offense 
  club is going to get joined onto possession team
  whichever team has possession find which players are moving
  
  
  
  1 .Explore the data to figure out what events are more or less rare
```{r}
#####
# 1 #
#####

#Filter out the na values from the data so we can figure out the frequency of 
#events

# Filter out NA values from the event column
wk1_events <- track1_df[!is.na(track1_df$event), ]

#Create a vector of axis labels that are more nicely formatted
events <- c("timeout_away"= "Time Out",
            "snap_direct"="Direct Snap",
            "shift"="Shift",
            "man_in_motion"="Man in Motion",
            "line_set"="Line Set",
            "huddle_start_offense"="Huddle Begin",
            "huddle_break_offense"="Huddle Break",
            "ball_snap"="Ball Snap")

# Create a bar plot of events (use geom_bar for categorical data)
event_hist <- wk1_events %>%
  ggplot(aes(x = event)) +
  geom_bar(fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Frequency of Events Before and During the Snap (Offense)",
    subtitle="Week 1 NFL Tracking Data (2022)",
    caption="Data from:NFL Big Data Bowl (2025) ",
    x = "Event",
    y = "Count"
  )+
  
  theme_bw()+
  
  #Add the better formatted labels to the plot
  scale_x_discrete(labels=events)+
  
  #Flip the axes so the labels are easier to read
  coord_flip()+
  
  theme(plot.title=element_text(hjust=0.5),
        plot.subtitle = element_text(hjust=0.5))

# Display the plot
event_hist

#Save the plot
ggsave("graphs/Images for paper/event_frequency_EDA.png", plot=event_hist,
       width=6.5, height=4)

#Shift is when the whole offense changes formation
#Man in motion is when one man moves across but the formation remains the same

```
Look up each of the event descriptions. Specifically look at what the man_in_motion event means 

Not finding anything useful in the kaggle description
  Depending on what man in motion means we might want to use that instead of our indicator variables because then we wouldn't have to worry about the na values
  
  
  
## 2 Data joining

Use the game_id and play_id to try and join the cleaned tracking data into the plays data.


```{r}
#####
# 2 #
#####

#Load in the plays data and glance at it to get an idea of how to join it to 
#the tracking data
play_glance <-head(plays_df, n=15)
track_glance <-head(track1_df)


#Try the simple approach of joining the week 1 tracking data with the 
#plays data by gameID and playID

# Perform an inner join on gameId and playId
merged_data <- track1_df %>%
  inner_join(plays_df, by = c("gameId", "playId"))

#Take a look at the resulting dataset and see if it looks right.
head(merged_data)


#Ok that seems to work super well. Have chanlder take a look at it to make sure
#nothing is being lost or if I'm missing something obvious. 

#Looking ahead not sure if it will be advantageous to add all of the tracking data into the plays data




```

So we are able to join the tracking data to the plays data. Now look at the code chandler used to do some cleaning and create our indicator variables.

Once I can get those joined I can find out how frequent runs vs. passes are

## 3. Run Chandlers code for cleaning and such and then try to merge the datasets

```{r}


```


## 4. Create indicator varaibles for run/pass and determine which is more common

```{r}
#####
# 4 #
#####

# # Create a new data frame by filtering out rows where dropbackType is NA or UNKNOWN
# filtered_data <- plays_df %>%
#   filter(!is.na(dropbackType) & dropbackType != "UNKNOWN")
# 
# # Create the indicator variable
# 
# #Run if -designed run or qb_sneak
# #Pass if - traditional, scramble, or contains "ROLLOUT"
# 
# filtered_data <- filtered_data %>%
#   mutate(run_pass_ind = case_when(
#     dropbackType %in% c("TRADITIONAL", "SCRAMBLE") | str_detect(dropbackType, "ROLLOUT") ~ 0,
#     dropbackType %in% c("QB_SNEAK", "DESIGNED_RUN") ~ 1,
#     TRUE ~ NA_real_ #for safety
#   ))
# 
# # Check the first few rows to ensure it worked
# head(filtered_data)
# 
# 
# #Plot the frequency of each type run vs. pass
# run_pass_plot <- filtered_data %>%
#   ggplot( aes(x = factor(run_pass_ind), fill = factor(run_pass_ind))) +
#   geom_bar() +
#   scale_x_discrete(labels = c("0" = "Pass", "1" = "Run")) +
#   labs(
#     x = "Play Type",
#     y = "Frequency",
#     title = "Frequency of Run and Pass Plays",
#     fill = "Play Type"
#   ) +
#   theme_minimal()
# 
# #Display the plot
# run_pass_plot




# Treat NA values as "DESIGNED_RUN" and UNKNOWN values as "OTHER"
filtered_data <- plays_df %>%
  mutate(dropbackType = ifelse(is.na(dropbackType) | dropbackType == "UNKNOWN", "DESIGNED_RUN", dropbackType))

# Create the indicator variable
# Run if - DESIGNED_RUN or QB_SNEAK
# Pass if - TRADITIONAL, SCRAMBLE, or contains "ROLLOUT"

filtered_data <- filtered_data %>%
  mutate(run_pass_ind = case_when(
    dropbackType %in% c("TRADITIONAL", "SCRAMBLE") | str_detect(dropbackType, "ROLLOUT") ~ 0,  
    # Pass
    dropbackType %in% c("QB_SNEAK", "DESIGNED_RUN") ~ 1,                                      
    # Run
    TRUE ~ NA_real_                                                                           
    # Safety fallback
  ))


# Check the first few rows to ensure it worked
head(filtered_data)

# Plot the frequency of each type run vs. pass
run_pass_plot <- filtered_data %>%
  ggplot(aes(x = factor(run_pass_ind), fill = factor(run_pass_ind))) +
  geom_bar() +
  scale_x_discrete(labels = c("0" = "Pass", "1" = "Run")) +
  theme_bw()+
  labs(
    x = "Play Type",
    y = "Frequency",
    title = "Frequency of Run and Pass Plays",
    subtitle= "2022 NFL Season (Week 1-9)",
    caption="Data from: NFL Big Data Bowl (2025)",
    fill = "Play Type"
  ) +
  theme(legend.position = "none",
        plot.title=element_text(hjust=0.5),
        plot.subtitle= element_text(hjust=0.5))

# Display the plot
run_pass_plot

#Save the plot
ggsave("graphs/Images for paper/play_type_frequency_EDA.png",
       plot=run_pass_plot, width=6.5, height=4)
```
That frequency seems really suspsicious. Try doing the same thing for all dropback types and see what their frequencies are

## 5. Investigate the frequency of each dropback type

```{r}
#####
# 5 #
#####

#Create a histogram showing the frequency of each kind of dropback type

# Assign "OTHER" to rows where dropbackType is NA
other_df <- plays_df %>%
  mutate(dropbackType = ifelse(is.na(dropbackType), "OTHER", dropbackType))


#Plot the frequency of each dropbackType
drop_type_plot <- other_df %>%
  ggplot( aes(x = factor(dropbackType), fill = factor(dropbackType))) +
  geom_bar() +
  labs(
    x = "Dropback Type",
    y = "Frequency",
    title = "Frequency of Dropback Types",
    fill = "Play Type"
  ) +
  theme(axis.text.x=element_text(angle=90))
  theme_minimal()

#Display the plot
drop_type_plot
```
From the plot I think we misunderstood the meaning of the TRADITIONAL drop back type. It must apply to several kinds of plays since it is outrageously more common than anything else. 

Look on the kaggle description to see if anything more specific is noted about the type.


UPDATE: ITs because the NA values are likely runs in actuality so removing them inflated the proportion of pass plays.



## 6. Try enconding NA values as OTHER and see what that does to the plot

```{r}

```

## 7. Need to find out which player on the offense is moving during presnap movement

If the man_in_motion indicator in the plays data is what we are looking for we can combine the tracking data to the play data with the same ID and try to figure out which player is in motion on the line 

Try this out using the merged data that I created in 2.



## 7 Try to figure out a way to figure out which player on the offense is moving

I'm hoping that I can find the man in motion and shift events to isolate plays that are relevant and then use some boundary to figure out which player is responsible for the movement 

```{r}
#####
# 7 #
#####

#Create dummy data frames for experimenting 
df7_plays <- plays_df
track1_df7 <- track1_df

#Merge the plays and tracking data
merged_df7 <- track1_df7 %>%
  inner_join(df7_plays, by = c("gameId", "playId"))

#Identify the relevant play ID's
   # i.e. plays that have a shift or man_in_motion event at some point during
   # the play.
relevant_play_ids <- merged_df7 %>%
  filter(event %in% c("man_in_motion", "shift")) %>%
  pull(playId)

#Create a data frame with just the relevant plays
filtered_df7 <- merged_df7 %>%
  filter(playId %in% relevant_play_ids)

#Calculate the average distance each player moved during the play
    #If all players have some degree of small noisy movement a simple sum might
    #be misleading since longer pre-snap times would artificially inflate the
    #distance traveled.

# Calculate the mean distance for each player for each play
mean_distance_df <- filtered_df7 %>%
  group_by(gameId, playId, displayName) %>%
  mutate(mean_dis = mean(dis, na.rm = TRUE)) %>%
  ungroup()

# View the results
head(mean_distance_df)

#Remove the football display name as that is not a player
mean_distance_df <- mean_distance_df %>%
  filter(displayName !="football")

#Create a vector of events that we are most interested in 
#Exclude:
#huddle_break_offense- not interested
#snap_direct- only 4 instances total and not sure what it means
#line_set- not interested in the motion of the offense getting to the line
valuable_events <- c("ball_snap", "man_in_motion", "shift")

#Plot the distribution of average distance traveled by event
mean_dist_plot <- mean_distance_df %>%
  #Play with some cutoff values to make sure that noisy shifting around isn't
  #included.
  filter(event %in% valuable_events, mean_dis >0.15) %>%
  ggplot()+
  geom_histogram(aes(x=mean_dis))+
  facet_wrap(~event)

#Display the plot
mean_dist_plot

#Think about diff cutoff for each
#keep the peak for ball_snap because only one person should be moving
#shift is slower so we would have a different cutoff
#possible to universal cutoff of 0.05



#Now try to create a new column that displays the names of all the players who
#were moving during that play 
high_dis_players <- mean_distance_df %>%
  filter(mean_dis >0.15)

#Add a column to the filtered data with the names of the players that were
#moving more than the cutoff on average

#something went wrong because it went up 1.6M observations so come back to fix
#this

# filtered_df7 <- filtered_df7 %>%
#   left_join(high_dis_players %>%
#               select(playId, gameId,displayName),
#             by=c("gameId", "playId","displayName")) %>%
#    mutate(high_mean_dis = ifelse(!is.na(displayName), displayName, NA_character_))
```

Chandler is going to look at movmement at the snap and movement after line set but before the snap.
Then we will remove 


## 8 Movement at ball snap

We are only really interested movement that occurs after the line is set. That is we are interested in any last second movement that occur before the snap. We want to use the frameId to identify when the line is set. Then we'll filter out the irrelevant data.

Note: The ball itself is included in the tracking data and is denoted with nflId=NA. Remove the NA values as we are only interested in players moving not the ball.

This process was done in the data cleaning file. Open the file and take a look at the filtered data.

```{r}

#Read in the master data set
line_set_df <- read.csv("data/tracking_line_set_frames.csv")

head(line_set_df)
```

which position moved, how far did they move (distance, average speed), and were they still moving at the time of the snap (), could do total distance

indicator of formation change
multiple people move and how far they moved------0.1 yds/0.1s

#Variables Final:


 play- plays 
 
, run vs. pass- plays data in chandlers EDA
, indicator for change of formation- tracking data

dummvars for who moved- use nflID to get the position from players data

distance_traveled (by position),- use first and last position to get euclidian distance

runpass lag,
epa lag,
offensve formation,
receiver alignment,
adverse weather indicator (just note that we would like to include this but its not available in the data provided)

success 50% of yards on 1st 70% on 2nd, 100% on 3rd and 4th down


```{r}
player_play_df <- read.csv("data/player_play.csv")

colnames(player_play_df)

players_df <- read.csv("data/players.csv")

colnames(players_df)

colnames(plays_df)
```

Variables:

run pass indicator from plays data

Change of formation: player_play data  (if there is a shift)
  shifts are counted as motions but motions arent counted as shifts
  
preSnapHomeTeamWinProbability- 
presnapVisitor team win prob- games.csv contains home away

home away indicator

Offensive formation: 2 columns in the plays offensive formation and receiver alignment




Run pass indicator:(rp_ind)
  -plays
  
Change of Formation:(change_form)
  -Use shift from player_play
  -if there was a shift
  
Pre-snap Home Team Win Probability: (preSnapVisitorTeamWinProbability):
  -
  
Offensive Formation: (offenseFormation)
  -
  
Expected Points Added Lag: (epa_lag)
  -
Receiver Alignment:
  -
  
Player movement:

shift_since_lineset:



-

Motion at ball snap:(inMotionAtBallSnap)

Motion since line set: (motionSinceLineset)
  -true if inmotion true
  -if true inmotion at snap not necesarily true
  




join all players
separate out wr, te, etc
join them back together into each row

-



```{r}
#Create a dummy dataframe with the columns we want to include

final_df_rd <- c("gameId", "playId", "rpInd", "offenseFormation", "epaLag",
                 "wrMotion", "teMotion","rbMotion",
                 "wrAtSnap", "teAtSnap", "rbAtSnap", "shiftSinceLineset",
                 "receiverAlignment", "currWinPercentage", "changeOfFormation","teamId",
                 "secondsLeftInHalf")




final_df_rd
```
Provided:

gameID, playID, offenseFormation, shiftSinceLineset, receiverAlginment,
presnapHomewinprob,  presnap Visitor, TeamId


To Calculate:


  Completed:
  -rpInd, epaLag, wrMotion, all the motions, 
  
  
  
  
  
first 6 weeks training
last 3 weeks testing


1st predicts 2nd
1,2 predict 3rd
1,2,3 predict 4
etc.
  
  



  
MOdel

lasso on whole thing, drop anythign at 0, logistic regression


lasso into logistic-Justin
random forrest-Gabe
knn- Chandler
svm- whoever gets there first




  











