---
title: "NFL Modeling - C"
author: "Chandler Grote"
date: "2024-11-22"
output: pdf_document
---


```{r setup, include=FALSE}
#Load the appropriate libraries:
library(ggplot2)
library(tidymodels)
library(tidyr)
library(tidyverse)
library(glmnet)
library(kknn)
library(caret)
```

# Modeling 

In this document we will develop a statistical model to predict whether the offense will execute a run or a pass play. The model will include the following variables


-game_id: Unique identifier for the game

-play_id: Unique identifier for a specific play

-possession_team: Identifies the offensive team

-week: Identifies which week of the season the game was in.

-run_pass: Indicates if the play was a run of a pass (response variable)

-down: Identifies which down the offensive team was on for a particular play.

-yards_to_go: The number of yards till a first down.

-yards_to_endzone: The number of yeards till the opponents endzone.

-redzone: Indicates whether the offense is within 20 yards of the opponents endzone.

-sec_in_half: The number of seconds remaining on the play clock in the current half.

-offense_formation: Categorical variable indicating what formation the offense uses.

-receiver_alingment: Enumerated as 0x0, 1x0, 1x1, 2x0, 2x1, 2x2, 3x0, 3x1, 3x2 (text)

-epa_lag: Serial correllation of expected points added.

-curr_win_percentage: Win percentage for the current offensive team.

-form_change: Uses shifts to determine if the offense changed formation before the ball was snapped.

-wr_motion: Indicates if a wide receiver moved after the line was set.

-te_motion: Indicates if a tight end moved after the line was set.

-rb_motion: Indicates if a running back moved after the line was set.

-wr_atsnap: Indicates if a wide receiver was in motion at the time of the snap.

-te_atsnap: Indicates if a tight end was in motion at the time of the snap.

-rb_atsnap: Indicates if a running back was in motion at the time of the snap.

# Modeling

## Cleaning the data frame

```{r data-frame-cleaning}
nfl_df <- read.csv("data/modeling_data_frame.csv")

# define unwanted columns to make modeling formula easier
unwanted_columns <- c("game_id", "play_id") 

# clean the df to remove unwanted columns and any rows containing NA values
df_clean <- nfl_df %>% select(-all_of(unwanted_columns)) %>% drop_na()
```

## KNN modeling test for one team

Define getting the bye week function
```{r}
get_bye_week <- function(weeks){
  # this gets the week number that is missing from the given weeks
  bye_week <- seq(9)[!(seq(9) %in% weeks)]
  
  # return NA if all weeks of data are present
  if(identical(bye_week, integer(0))){
    return(NA)
  }else{
    return(bye_week)
  }
}

test_data_for_augment <- function(model, data){
  
}
```


```{r knn-one-team}
set.seed(445)

model_info <- list("nfl_team" = unique(df_clean$possession_team),
                   "model" = as.list(seq(32)),
                   "bye_week" = as.list(seq(32)),
                   "train_fit" = as.list(seq(32)),
                   "test_fit" = as.list(seq(32)),
                   "conf_mat" = as.list(seq(32)))

# initialize a large grid of k vlaues for CV
grid_large <- tibble(neighbors = seq(0, 100, by = 5)[-1])

for(i in c(1:32)){
  team <- model_info$nfl_team[i]
  
  # if(team %in% break_teams){next}
  
  print(paste("Working on team", team))
  
  team_df <- df_clean %>% 
    filter(possession_team == team) %>% 
    mutate(run_pass = as.factor(run_pass))
  
  weeks <- sort(unique(team_df$week))
  
  model_info$bye_week[[i]] <- get_bye_week(weeks)
  
  # training is everything but the last 3 weeks in the data for that team.
  # this could be more for some teams if they did not have a bye week in first 9 weeks
  training_df <- team_df %>% 
    filter(week <= weeks[(length(weeks)-3)]) %>% select(-possession_team, -week)
  # testing data is all weeks not in training
  testing_df <- team_df %>% 
    filter(week > weeks[(length(weeks)-3)]) %>% select(-possession_team, -week)
  
  knn_tune_spec <- nearest_neighbor(mode = "classification", neighbors = tune("neighbors"))
  
  train_cv <- vfold_cv(training_df, v = 10)
  
  knn_err <- knn_tune_spec %>%
    tune_grid(run_pass ~ . , resamples = train_cv, grid = grid_large) %>%
    collect_metrics() %>%
    dplyr::filter(.metric == "accuracy") %>%
    dplyr::mutate(error = 1 - mean)
  
  # plot if necessary
  # ggplot(knn_err) + geom_line(aes(x = neighbors, y = error))
  
  # get minimum k value from first CV
  knn_min_range <- knn_err$neighbors[knn_err$error == min(knn_err$error)]
  print(paste("Min K is near", knn_min_range))
  
  # build a narrowing k range around the minimum found
  grid_seq <- seq(knn_min_range[1] - 4, knn_min_range[1] + 4, by = 1)
  grid_small <-  tibble(neighbors = grid_seq)
  
  # re-run to find overall minimum
  knn_err <- knn_tune_spec %>%
    tune_grid(run_pass ~ . , resamples = train_cv, grid = grid_small) %>%
    collect_metrics() %>%
    dplyr::filter(.metric == "accuracy") %>%
    dplyr::mutate(error = 1 - mean)
  
  
  # plot if necessary
  # ggplot(knn_err) + geom_line(aes(x = neighbors, y = error))
  
  # get new min k
  min_k <- min(knn_err$neighbors[knn_err$error == min(knn_err$error)])
  
  print(paste("Overall Min K is", min_k))
  
  # fit the model with k value found in CV
  knn_fit <- nearest_neighbor(mode = "classification", neighbors = min_k) %>% 
    fit(data = training_df, run_pass ~ .)
  model_info$model[[i]] <- knn_fit
  
  # get the fit from the tidy model
  train_fit <- knn_fit %>% pluck("fit")
  model_info$train_fit[[i]] <- train_fit
  
  # check test data against the model
  test_predict <- tryCatch({knn_fit %>% augment(testing_df)}, 
                           error = function(cond){
                             print(conditionMessage(cond)) 
                             NA})
  
  if(length(is.na(test_predict)) > 1){
    model_info$test_fit[[i]] <- test_predict
    
    # test_predict %>% conf_mat(truth = run_pass, estimate = .pred_class)
    
    test_conf_mat <- confusionMatrix(test_predict$.pred_class, test_predict$run_pass)
    model_info$conf_mat[[i]] <- test_conf_mat
  }else{
    model_info$test_fit[[i]] <- paste(team, "errored out")
    
    model_info$conf_mat[[i]] <- paste(team, "errored out")
  }
  
}
```


```{r}
set.seed(445)

model_info <- list("nfl_team" = unique(df_clean$possession_team),
                   "model" = as.list(seq(32)),
                   "bye_week" = as.list(seq(32)),
                   "train_fit" = as.list(seq(32)),
                   "test_fit" = as.list(seq(32)),
                   "conf_mat" = as.list(seq(32)))

# initialize a large grid of k vlaues for CV
grid_large <- tibble(neighbors = seq(0, 100, by = 5)[-1])

for(i in c(1:32)){
  team <- model_info$nfl_team[i]
  
  # if(team %in% break_teams){next}
  
  print(paste("Working on team", team))
  
  team_df <- df_clean %>% 
    filter(possession_team == team) %>% 
    mutate(run_pass = as.factor(run_pass))
  
  weeks <- sort(unique(team_df$week))
  
  model_info$bye_week[[i]] <- get_bye_week(weeks)
  
  # training is going to be a 60% split for each team
  # This is an attempt to avoid errors
  team_split <- team_df %>% initial_split(prop = 0.5, strata = "run_pass")
  team_training <- team_split %>% training() %>% select(-possession_team, -week)
  team_testing <- team_split %>% testing() %>% select(-possession_team, -week)
  
  # training_df <- team_df %>% 
  #   filter(week <= weeks[(length(weeks)-3)]) %>% select(-possession_team, -week)
  # # testing data is all weeks not in training
  # testing_df <- team_df %>% 
  #   filter(week > weeks[(length(weeks)-3)]) %>% select(-possession_team, -week)
  
  knn_tune_spec <- nearest_neighbor(mode = "classification", neighbors = tune("neighbors"))
  
  train_cv <- vfold_cv(team_training, v = 10)
  
  knn_err <- knn_tune_spec %>%
    tune_grid(run_pass ~ . , resamples = train_cv, grid = grid_large) %>%
    collect_metrics() %>%
    dplyr::filter(.metric == "accuracy") %>%
    dplyr::mutate(error = 1 - mean)
  
  # plot if necessary
  # ggplot(knn_err) + geom_line(aes(x = neighbors, y = error))
  
  # get minimum k value from first CV
  knn_min_range <- knn_err$neighbors[knn_err$error == min(knn_err$error)]
  print(paste("Min K is near", knn_min_range))
  
  # build a narrowing k range around the minimum found
  grid_seq <- seq(knn_min_range[1] - 4, knn_min_range[1] + 4, by = 1)
  grid_small <-  tibble(neighbors = grid_seq)
  
  # re-run to find overall minimum
  knn_err <- knn_tune_spec %>%
    tune_grid(run_pass ~ . , resamples = train_cv, grid = grid_small) %>%
    collect_metrics() %>%
    dplyr::filter(.metric == "accuracy") %>%
    dplyr::mutate(error = 1 - mean)
  
  
  # plot if necessary
  # ggplot(knn_err) + geom_line(aes(x = neighbors, y = error))
  
  # get new min k
  min_k <- min(knn_err$neighbors[knn_err$error == min(knn_err$error)])
  
  print(paste("Overall Min K is", min_k))
  
  # fit the model with k value found in CV
  knn_fit <- nearest_neighbor(mode = "classification", neighbors = min_k) %>% 
    fit(data = team_training, run_pass ~ .)
  model_info$model[[i]] <- knn_fit
  
  # get the fit from the tidy model
  train_fit <- knn_fit %>% pluck("fit")
  model_info$train_fit[[i]] <- train_fit
  
  # check test data against the model
  test_predict <- tryCatch({knn_fit %>% augment(team_testing)}, 
                           error = function(cond){
                             print(conditionMessage(cond)) 
                             NA})
  
  if(length(is.na(test_predict)) > 1){
    model_info$test_fit[[i]] <- test_predict
    
    # test_predict %>% conf_mat(truth = run_pass, estimate = .pred_class)
    
    test_conf_mat <- confusionMatrix(test_predict$.pred_class, test_predict$run_pass)
    model_info$conf_mat[[i]] <- test_conf_mat
  }else{
    model_info$test_fit[[i]] <- paste(team, "errored out")
    
    model_info$conf_mat[[i]] <- paste(team, "errored out")
  }
  
}
```




## SVM modeling

```{r}

```


# Copied from Justin's code
```{r}

#Load in the data set
nfl_df <- read.csv("data/modeling_data_frame.csv")



#Make the run pass variable a factor
nfl_df$run_pass <- as.factor(nfl_df$run_pass)

#Convert the boolean columns into numeric ones
bool_cols <- c("redzone", "wr_motion", "te_motion", "rb_motion",
               "wr_atsnap", "te_atsnap", "rb_atsnap")
nfl_df[bool_cols] <- lapply(nfl_df[bool_cols], as.numeric)

#Remove NA values
nfl_df <- nfl_df %>%
  drop_na()

#Use a lasso model to do varaible selection.

#Create a sequence of lambda values to try
lambda <- lambda <- 10^seq(-2, 10, length.out = 100)

#Create a recipe for the lasso model and normalize the predictors.
lasso_rec <- recipe(run_pass~., data=nfl_df) %>%
  
  #Remove unnecessary variables
  step_rm(game_id, play_id, possession_team,week) %>%
  
  step_novel(all_nominal_predictors()) %>%
  
  #Fix error for missing values by assigning them with unknown
  step_unknown(all_nominal_predictors()) %>%
  
  #Fix error where some variables are categorical and not numeric
  step_dummy(all_nominal_predictors()) %>%
  
  #Fix error from some of them having zero variance
  step_zv(all_predictors())%>%
  
  #Normalize the numeric predictors to have mean=0 and sd=1
  step_normalize(all_numeric_predictors())

#Create an empty data frame to store our lasso estimates
lasso_ests <- data.frame()

#Fit a lasso model for each lambda in our grid
for(lam in lambda){
  lasso_spec <- logistic_reg(mixture=1, penalty=lam)%>%
    set_mode("classification") %>%
    set_engine("glmnet")
  
  #Fit the model
  workflow() %>%
    add_model(lasso_spec) %>%
    add_recipe(lasso_rec) %>%
    fit(nfl_df) %>%
    tidy() %>%
    bind_rows(lasso_ests)-> lasso_ests
}

  #Create a line plot of 
lasso_ests |>
  filter(term != "(Intercept)") |>
  ggplot() +
  geom_line(aes(penalty, estimate, group = term, colour = term)) +
  scale_x_continuous(limits = c(0.01, .2), breaks = seq(0, 0.15, 0.025)) +
  coord_trans(x = "log10")+
  labs(title = "NFL Run-Pass Relation with Predictors",
       subtitle = "LASSO Implementation for different lambda values",
       x = "Lambda Penalty", y = "Coefficient Estimate (beta_hat)",
       caption = "data from NFL Big Data Bowl Competition") +
  #The legend is so large it obstructs the graph. You can comment out this line
  # and view the graph in a new window to get a better idea. 
  theme(legend.position="none")
```


