---
title: "NFL EDA - J"
author: "Justin Eldridge"
date: "2024-11-13"
output: html_document
---

# EDA for Justin to work on to avoid repo desync

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(broom)
library(GGally)
library(tidyr)
library(tidymodels) 
library(kknn)
library(tidyverse)
library(ggh4x)
```

#Try to select a model to predict pass vs. run

```{r}


#load in the tracking data for week 1
track1_df <- read.csv("data/presnap_events_week_1.csv")

#Explor the variables available to us
colnames(track1_df)

#Load in the plays and games data to explore the variables
plays_df <- read.csv("data/plays.csv")

colnames(plays_df)
```

## Brainstorm on variables to include

playid- to see what the previous play was to do a lag (i.e. if the previous run failed they are more liekly to do a pass)

time_remaining

yards remaining

score

maybe indicate if there was pre-snap movement 
  can create magnitude vectors and take a difference for each one
  use x coordinates
  
  orientation 360 where they ar facing 
  
distance direction and orientation for pre-snap movement 

If over a certain number indicate that pre-snap movement took place

want the sum of the distances traveled in each event before the snap

create our own distance column because they could still be moving even if their was no event. So when we removed the na values we removed some of the distances traveled before the snap

possibly filter out play with distance <0.1

Need to change the data cleaning file

to do lag on play we might have to sort by game ID, play id

red_zone 

possibly come back and make a function to calculate the teams record 

should exclude plays with a penalty would introduce too much error

exclude qbkneels, 





Indicator for significant pre-snap movement
time remaining
yards remaining
score
red-zone
play_id
game_id
play-concept lag
See if the teams current record is available
prensnap home and visitor score from the plays data
designed sneaks included don't include forced sneaks might be captured in run
runpassoption

either have to exclude rpo (don't happen often enough), or can include in prediction model as either run pass or rpo
  typcially fairly rare
  
  svm wouldnt be good for 3 classifiers, best for binary classifier
  
  or consider RPO vs run run 1st pass 2nd if the run isnt there then throw i
  so we would consider this a run play 
  Consider and rpo to be a run play
  if we do this just call them runs and then we could still use svm for the binary run pass classifier
  
  every rpo has a primary run concept pff thingy so we could use that column to classifiy 
  
  Break the data down into individual drives otherwise an epa lag will consider the other teams score in the new drive. 
  


rpo the qb chooses as the play unfolds whether hes going to pass or run



pass_length to try and get and indicator varialbe
could use rush_location_type

runconcept primary sometiems show as having one even if its a pass
  bc of designed rollout in dropback type have a run concept associated for some reason
  
  can use droback type as better indicator of pass
    -traditional scrabmle for passes, desgined rollouts-passes, 
    elminiate unknown dropback types
    
calculate yard to endzone because data fails to account for the 10 yards in the endzone(based on yard number)
-dont use yardstogo



2nd draft of variables:

  time remaining
  score
  yards remaining
  epa lag (break down into drives so lag works properly)
  run/pass indicator (include runpass options as runs so we can do svm,
  designed qb sneaks as runs scrambles as passes)
  play-concept lag
  red_zone
  game_id
  play_id
  indicator var for significant pre-snap movement
  pre-snap home and visitor score
  
  --they have var with game clock which resets every quarter
  might be able to make a function that takes game clock and gives us the amount of seconds remaining until end of the half
  
  
  
  to do:
  
  some eda
  
  how man runs vs. passes
  how often certain events occur look for rare events
  how many rpos end up as passes
  
  
  do svm for offense
  
  might do random forrest for defense
  two binary classifiers create 4 possile situtations
  
  
  data cleaning:
  
  -- make a function to convert the game clock into seconds remaining in the half.
  --play data, time on clock of play in minutes seconds.
  
  work on cleaning for the presnap events-
  use game id and play id to join the data, to attach cleaned data to the plays data
  
  neeed out of presnap, which player is moving on the offense 
  club is going to get joined onto possession team
  whichever team has possession find which players are moving
  
  
  
  1 .Explore the data to figure out what events are more or less rare
```{r}
#####
# 1 #
#####

#Filter out the na values from the data so we can figure out the frequency of 
#events

# Filter out NA values from the event column
wk1_events <- track1_df[!is.na(track1_df$event), ]

# Create a bar plot of events (use geom_bar for categorical data)
event_hist <- wk1_events %>%
  ggplot(aes(x = event)) +
  geom_bar(fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Frequency of Events",
    x = "Event",
    y = "Count"
  )+
  #Rotate the axis labels to make them more readable
  theme(axis.text.x= element_text(angle=45))

# Display the plot
event_hist

```
Look up each of the event descriptions. Specifically look at what the man_in_motion event means 

Not finding anything useful in the kaggle description
  Depending on what man in motion means we might want to use that instead of our indicator variables because then we wouldn't have to worry about the na values
  
  
  
## 2 Data joining

Use the game_id and play_id to try and join the cleaned tracking data into the plays data.


```{r}
#####
# 2 #
#####

#Load in the plays data and glance at it to get an idea of how to join it to 
#the tracking data

play_glance <-head(plays_df, n=15)

track_glance <-head(track1_df)


#Try the simple approach of joining the week 1 tracking data with the 
#plays data by gameID and playID


library(dplyr)

# Perform an inner join on gameId and playId
merged_data <- track1_df %>%
  inner_join(plays_df, by = c("gameId", "playId"))

#Take a look at the resulting dataset and see if it looks right.
head(merged_data)


#Ok that seems to work super well. Have chanlder take a look at it to make sure
#nothing is being lost or if I'm missing something obvious. 

#Looking ahead not sure if it will be advantageous to add all of the tracking data into the plays data




```

So we are able to join the tracking data to the plays data. Now look at the code chandler used to do some cleaning and create our indicator variables.

Once I can get those joined I can find out how frequent runs vs. passes are

## 3. Run Chandlers code for cleaning and such and then try to merge the datasets

```{r}
#####
# 3 #
#####

plays <- read.csv("data/plays.csv")

presnap_1 <- read.csv("data/presnap_events_week_1.csv")

week_1_plays <- plays %>% filter(gameId %in% presnap_1$gameId, 
                                 playNullifiedByPenalty == "N", 
                                 qbKneel == 0,
                                 (dropbackType != "UNKNOWN" | is.na(dropbackType))) %>%
  arrange(gameId, playId) %>%
  mutate(run_pass = case_when(pff_runPassOption == 1 ~ "run",
                              is.na(dropbackType) ~ "run",
                              str_detect(dropbackType, "RUN|SNEAK") ~ "run",
                              T ~ "pass"),
         yards_to_endzone = ifelse(possessionTeam == yardlineSide, abs(yardlineNumber - 100), yardlineNumber),
         redzone = ifelse(yards_to_endzone <= 20, T, F),
         half = ifelse(quarter <= 2, 1, 2),
         sec_in_half = ((as.numeric(str_extract(gameClock, "[:digit:]+(?=:)")) * 60) + as.numeric(str_extract(gameClock, "(?<=:)[:digit:]+"))) + ifelse(quarter %in% c(1, 3), 900, 0))

week_1_plays$new_half <- week_1_plays$half != c(0, week_1_plays$half[-nrow(week_1_plays)])
week_1_plays$new_game <- week_1_plays$gameId != c(0, week_1_plays$gameId[-nrow(week_1_plays)])
week_1_plays$new_possession <- (week_1_plays$possessionTeam != c(0, week_1_plays$possessionTeam[-nrow(week_1_plays)])) | week_1_plays$new_half

# set first row of variables
teams <- c(week_1_plays$possessionTeam[1], week_1_plays$defensiveTeam[1])
drive <- c(1, 0)
names(drive) <- teams
week_1_plays$drive <- c(1, rep(0, nrow(week_1_plays) - 1))

# loop through rows 2 to end of df
for(i in c(2:nrow(week_1_plays))){
  if(week_1_plays$new_game[i]){
    teams <- c(week_1_plays$possessionTeam[i], week_1_plays$defensiveTeam[i])
    drive <- c(1, 0)
    names(drive) <- teams
  }
  else if(week_1_plays$new_possession[i]){
    drive[week_1_plays$possessionTeam[i]] <- drive[week_1_plays$possessionTeam[i]] + 1
  }
  week_1_plays$drive[i] <- drive[week_1_plays$possessionTeam[i]]
}

# week_1_plays[,c("possessionTeam", "new_possession", "drive", "new_half")]

```


## 4. Create indicator varaibles for run/pass and determine which is more common

```{r}
#####
# 4 #
#####

# Create a new data frame by filtering out rows where dropbackType is NA or UNKNOWN
filtered_data <- plays_df %>%
  filter(!is.na(dropbackType) & dropbackType != "UNKNOWN")

# Create the indicator variable

#Run if -designed run or qb_sneak
#Pass if - traditional, scramble, or contains "ROLLOUT"

filtered_data <- filtered_data %>%
  mutate(run_pass_ind = case_when(
    dropbackType %in% c("TRADITIONAL", "SCRAMBLE") | str_detect(dropbackType, "ROLLOUT") ~ 0,
    dropbackType %in% c("QB_SNEAK", "DESIGNED_RUN") ~ 1,
    TRUE ~ NA_real_ #for safety
  ))

# Check the first few rows to ensure it worked
head(filtered_data)


#Plot the frequency of each type run vs. pass
run_pass_plot <- filtered_data %>%
  ggplot( aes(x = factor(run_pass_ind), fill = factor(run_pass_ind))) +
  geom_bar() +
  scale_x_discrete(labels = c("0" = "Pass", "1" = "Run")) +
  labs(
    x = "Play Type",
    y = "Frequency",
    title = "Frequency of Run and Pass Plays",
    fill = "Play Type"
  ) +
  theme_minimal()

#Display the plot
run_pass_plot

library(ggplot2)

# Assuming your dataset is named filtered_data
ggplot(filtered_data, aes(x = factor(run_pass_indicator), fill = factor(run_pass_indicator))) +
  geom_bar() +
  scale_x_discrete(labels = c("0" = "Pass", "1" = "Run")) +
  labs(
    x = "Play Type",
    y = "Frequency",
    title = "Frequency of Run and Pass Plays",
    fill = "Play Type"
  ) +
  theme_minimal()


```
That frequency seems really suspsicious. Try doing the same thing for all dropback types and see what their frequencies are

## 5. Investigate the frequency of each dropback type

```{r}
#####
# 5 #
#####

#Create a histogram showing the frequency of each kind of dropback type


#Plot the frequency of each dropbackType
drop_type_plot <- filtered_data %>%
  ggplot( aes(x = factor(dropbackType), fill = factor(dropbackType))) +
  geom_bar() +
  labs(
    x = "Dropback Type",
    y = "Frequency",
    title = "Frequency of Dropback Types",
    fill = "Play Type"
  ) +
  theme(axis.text.x=element_text(angle=90))
  theme_minimal()

#Display the plot
drop_type_plot
```
From the plot I think we misunderstood the meaning of the TRADITIONAL drop back type. It must apply to several kinds of plays since it is outrageously more common than anything else. 

Look on the kaggle description to see if anything more specific is noted about the type.




  
  
  











