---
title: "NFL Modeling - C"
author: "Chandler Grote"
date: "2024-11-22"
output: pdf_document
---


```{r setup, include=FALSE}
#Load the appropriate libraries:
library(ggplot2)
library(tidymodels)
library(tidyr)
library(tidyverse)
library(glmnet)
library(kknn)
```

# Modeling 

In this document we will develop a statistical model to predict whether the offense will execute a run or a pass play. The model will include the following variables


-game_id: Unique identifier for the game

-play_id: Unique identifier for a specific play

-possession_team: Identifies the offensive team

-week: Identifies which week of the season the game was in.

-run_pass: Indicates if the play was a run of a pass (response variable)

-down: Identifies which down the offensive team was on for a particular play.

-yards_to_go: The number of yards till a first down.

-yards_to_endzone: The number of yeards till the opponents endzone.

-redzone: Indicates whether the offense is within 20 yards of the opponents endzone.

-sec_in_half: The number of seconds remaining on the play clock in the current half.

-offense_formation: Categorical variable indicating what formation the offense uses.

-receiver_alingment: Enumerated as 0x0, 1x0, 1x1, 2x0, 2x1, 2x2, 3x0, 3x1, 3x2 (text)

-epa_lag: Serial correllation of expected points added.

-curr_win_percentage: Win percentage for the current offensive team.

-form_change: Uses shifts to determine if the offense changed formation before the ball was snapped.

-wr_motion: Indicates if a wide receiver moved after the line was set.

-te_motion: Indicates if a tight end moved after the line was set.

-rb_motion: Indicates if a running back moved after the line was set.

-wr_atsnap: Indicates if a wide receiver was in motion at the time of the snap.

-te_atsnap: Indicates if a tight end was in motion at the time of the snap.

-rb_atsnap: Indicates if a running back was in motion at the time of the snap.

# Modeling

## Cleaning the data frame

```{r data-frame-cleaning}
nfl_df <- read.csv("data/modeling_data_frame.csv")

unwanted_columns <- c("game_id", "play_id")

df_clean <- nfl_df %>% select(-all_of(unwanted_columns))
```

## KNN modeling test for one team

```{r knn-one-team}
set.seed(445)
team <- "DEN"

team_df <- df_clean %>% 
  filter(possession_team == team, !is.na(run_pass), !is.na(yards_to_endzone)) %>% 
  mutate(run_pass = as.factor(run_pass))

# start with using weeks 1-5 to predict week 6 and see how it goes
training_df <- team_df %>% 
  filter(week < 6) %>% select(-possession_team, -week)
testing_df <- team_df %>% 
  filter(week %in% c(6, 7, 8)) %>% select(-possession_team, -week)

grid_large <- tibble(neighbors = seq(0, 100, by = 5)[-1])

knn_tune_spec <- nearest_neighbor(mode = "classification", neighbors = tune("neighbors"))

train_cv <- vfold_cv(training_df, v = 10)

knn_err <- knn_tune_spec %>%
  tune_grid(run_pass ~ . , resamples = train_cv, grid = grid_large) %>%
  collect_metrics() %>%
  dplyr::filter(.metric == "accuracy") %>%
  dplyr::mutate(error = 1 - mean)

ggplot(knn_err) + geom_line(aes(x = neighbors, y = error))

knn_min_range <- knn_err$neighbors[knn_err$error == min(knn_err$error)]

grid_seq <- seq(knn_min_range[1] - 4, knn_min_range[1] + 4, by = 1)
grid_small <-  tibble(neighbors = grid_seq)

knn_err <- knn_tune_spec %>%
  tune_grid(run_pass ~ . , resamples = train_cv, grid = grid_small) %>%
  collect_metrics() %>%
  dplyr::filter(.metric == "accuracy") %>%
  dplyr::mutate(error = 1 - mean)

ggplot(knn_err) + geom_line(aes(x = neighbors, y = error))

min_k <- min(knn_err$neighbors[knn_err$error == min(knn_err$error)])

knn_fit <- nearest_neighbor(mode = "classification", neighbors = min_k) %>% 
  fit(data = training_df, run_pass ~ .)

train_fit <- knn_fit %>% pluck("fit")

test_predict <- knn_fit %>% augment(testing_df)

test_predict %>% conf_mat(truth = run_pass, estimate = .pred_class)

```



