---
title: "NFL-Modeling---J"
author: "Justin C. Eldridge"
date: "2024-11-22"
output: pdf_document
---





```{r setup, include=FALSE}
#Load the appropriate libraries:
library(ggplot2)
library(tidymodels)
library(tidyr)
library(tidyverse)

library(glmnet)

```

# Modeling 

  In this document we will develop a statistical model to predict whether the offense will execute a run or a pass play. The model will include the following variables
  
  
-game_id: Unique identifier for the game

-play_id: Unique identifier for a specific play

-possession_team: Identifies the offensive team

-week: Identifies which week of the season the game was in.

-run_pass: Indicates if the play was a run of a pass (response variable)

-down: Identifies which down the offensive team was on for a particular play.

-yards_to_go: The number of yards to the opponents endzone.

-redzone: Indicates whether the offense is within 20 yards of the opponents endzone.

-sec_in_half: The number of seconds remaining on the play clock in the current half.

-offense_formation: Categorical variable indicating what formation the offense uses.

-receiver_alingment: Enumerated as 0x0, 1x0, 1x1, 2x0, 2x1, 2x2, 3x0, 3x1, 3x2 (text)

-epa_lag: Serial correllation of expected points added.

-curr_win_percentage: Win percentage for the current offensive team.

-form_change: Uses shifts to determine if the offense changed formation before the ball was snapped.

-wr_motion: Indicates if a wide receiver moved after the line was set.

-te_motion: Indicates if a tight end moved after the line was set.

-rb_motion: Indicates if a running back moved after the line was set.

-wr_atsnap: Indicates if a wide receiver was in motion at the time of the snap.

-te_atsnap: Indicates if a tight end was in motion at the time of the snap.

-rb_atsnap: Indicates if a running back was in motion at the time of the snap.

# Model 1: Logistic (Control)

  Use LASSO to perform varaible selection and then fit a logistic regression model as a kind of control to compare our other models against.

```{r, logistic_model, cache=TRUE}

#Load in the data set
nfl_df <- read.csv("data/modeling_data_frame.csv")



#Make the run pass variable a factor
nfl_df$run_pass <- as.factor(nfl_df$run_pass)

#Convert the boolean columns into numeric ones
bool_cols <- c("redzone", "wr_motion", "te_motion", "rb_motion",
               "wr_atsnap", "te_atsnap", "rb_atsnap")
nfl_df[bool_cols] <- lapply(nfl_df[bool_cols], as.numeric)

#Remove NA values
nfl_df <- nfl_df %>%
  drop_na()

#Use a lasso model to do varaible selection.

#Create a sequence of lambda values to try
lambda <- lambda <- 10^seq(-2, 10, length.out = 100)

#Create a recipe for the lasso model and normalize the predictors.
lasso_rec <- recipe(run_pass~., data=nfl_df) %>%
  
  #Remove unnecessary variables
  step_rm(game_id, play_id, possession_team,week) %>%
  
  step_novel(all_nominal_predictors()) %>%
  
  #Fix error for missing values by assigning them with unknown
  step_unknown(all_nominal_predictors()) %>%
  
  #Fix error where some variables are categorical and not numeric
  step_dummy(all_nominal_predictors()) %>%
  
  #Fix error from some of them having zero variance
  step_zv(all_predictors())%>%
  
  #Normalize the numeric predictors to have mean=0 and sd=1
  step_normalize(all_numeric_predictors())

#Create an empty data frame to store our lasso estimates
lasso_ests <- data.frame()

#Fit a lasso model for each lambda in our grid
for(lam in lambda){
  lasso_spec <- logistic_reg(mixture=1, penalty=lam)%>%
    set_mode("classification") %>%
    set_engine("glmnet")
  
  #Fit the model
  workflow() %>%
    add_model(lasso_spec) %>%
    add_recipe(lasso_rec) %>%
    fit(nfl_df) %>%
    tidy() %>%
    bind_rows(lasso_ests)-> lasso_ests
}

  #Create a line plot of 
lasso_ests |>
  filter(term != "(Intercept)") |>
  ggplot() +
  geom_line(aes(penalty, estimate, group = term, colour = term)) +
  coord_trans(x = "log10")+
  #The legend is so large it obstructs the graph. You can comment out this line
  # and view the graph in a new window to get a better idea. 
  theme(legend.position="none")
```
```{r}

#Create the training and testing sets of data
  #Use the first 6 weeks of data for training 
  #Use the last 3 weeks of dat for testing

nfl_lasso_df <- nfl_df

nfl_lasso_df$run_pass <- as.factor(nfl_lasso_df$run_pass)

nfl_lasso_df[bool_cols] <- lapply(nfl_lasso_df[bool_cols], as.numeric)

nfl_train <- nfl_lasso_df %>%
  filter(week==c(1,2,3,4,5,6))

nfl_test <- nfl_lasso_df %>%
  filter(week==c(7,8,9))
  
  

  
```



```{r}

#Create a data frame for tuning lambda
tune_df <- data.frame(lambda=lambda)

#Perform 10-fold cross-validation to get an estimate of test MSE for each
#labmda in our grid
nfl_10fld_cv <- vfold_cv(nfl_df, v=10)

#Specify the model
lasso_spec <- logistic_reg(mixture = 1, penalty = tune("lambda")) %>%
  set_mode("classification") %>%
  set_engine("glmnet")

#Create the workflow object for lasso
lasso_workflow <-workflow() %>%
  add_model(lasso_spec) %>%
  add_recipe(lasso_rec) 

#Perform the cross validation
tune_results <- tune_grid( lasso_workflow,
                           resamples=nfl_10fld_cv,
                           #For some reason it works if I let the labmda values
                           #be created automatically but not when I try to use
                           #the labmda values in the tuning 
                           grid=50)

#Determine the best lambda value
best_lambda <- select_best(tune_results, metric="accuracy") 

#Create the lasso model with the optimum lambda
final_lasso_workflow <- finalize_workflow(lasso_workflow,  best_lambda)

final_lasso_model <- fit(final_lasso_workflow, data=nfl_train) -> final_lasso_fit

final_model <- extract_fit_parsnip(final_lasso_model)



#Convert the coefficients to a tidy data frame
selected_variables <- tidy(final_model)


#Generate predictions on the test dataset
lasso_predictions <- predict(final_lasso_model, nfl_test, type = "class") %>%
  bind_cols(nfl_test)

#Create a confusion matrix
confusion_mat <- conf_mat(
  data = lasso_predictions,
  truth = run_pass, 
  estimate = .pred_class
)

#Print the confusion matrix
print(confusion_mat)


```

```{r}
# Extract non-zero coefficient variables from the LASSO model
# selected_variables <- tidy(final_model) %>%
#   filter(term != "(Intercept)" & estimate != 0) %>%
#   pull(term)

selected_vars <-c("down", "yards_to_go", "yards_to_endzone", "sec_in_half",
                  "epa_lag", "curr_win_percentage", "form_change", "wr_motion",
                  "te_motion","rb_motion","wr_atsnap", "te_atsnap",
                  "offense_formation", "receiver_alignment")

#The test set has a level of offense_formation that is not in the training set

#Select the cagetorical variables with multiple levels
factor_vars <- c("offense_formation", "receiver_alignment")

#Set the levels in the test set to be the same as the levels in teh training set
for (var in factor_vars) {
  nfl_test[[var]] <- factor(nfl_test[[var]], levels = levels(nfl_train[[var]]))
}

#Create a formula for logistic regression using selected variables
logistic_formula <- as.formula(paste("run_pass ~",
                                     paste(selected_vars, collapse = " + ")))

#Fit the logistic regression model using the selected variables
final_logistic_model <- glm(
  formula = logistic_formula,
  data = nfl_train,
  family = binomial()
)

# Generate predictions on the test set
logistic_predictions_prob <- predict(final_logistic_model,
                                     newdata = nfl_test, type = "response")

#Convert probabilities to class labels based on a threshold of 0.5
logistic_predictions_class <- ifelse(logistic_predictions_prob > 0.5, "run", "pass")

#Add predictions to the test dataset
nfl_test <- nfl_test %>%
  mutate(predicted_run_pass = logistic_predictions_class)

#Create a confusion matrix
confusion_matrix <- table(
  Truth = nfl_test$run_pass,
  Prediction = nfl_test$predicted_run_pass
)

#Print the confusion matrix
print(confusion_matrix)

#Calculate the accuracy and error rates
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
error_rate <- 1 - accuracy

cat("Accuracy Rate:", accuracy, "\n")
cat("Error Rate:", error_rate, "\n")




```






