---
title: "Reproduce Logistic Model (LASSO)"
author: "Justin C. Eldridge"
date: "2024-12-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Load the appropriate libraries:
library(ggplot2)
library(tidymodels)
library(tidyr)
library(glmnet)
library(mltools)
library(caret)
library(nflplotR)
library(dplyr)
library(tidyverse)
```

# Logistic Model (LASSO)

This document contains all the code necessary to reproduce the logistic models along with the associated visualizations.

## Contents:
  1. Variables
  2. Lasso Model
  3. Tuning Lasso
  4. Logistic Regression Model (Entire Data Set)
  5. Logistic Regression Models by Team
  6. Visualizations
  

## 1. Variables:

  Using our background knowledge and drawing on our EDA we decided to include the following variables in our models.
  
  
  **_Game Characteristics:_**

- **`game_id`**: Unique identifier for the game
- **`play_id`**: Unique identifier for a specific play
- **`possession_team`**: Identifies the offensive team
- **`week`**: Identifies which week of the season the game was in
- **`run_pass`**: Indicates if the play was a run or a pass (response variable)
- **`down`**: Identifies which down the offensive team was on for a particular play
- **`yards_to_go`**: The number of yards needed to get a first down
- **`yards_to_endzone`**: The number of yards to the opponent's endzone
- **`redzone`**: Indicates whether the offense is within 20 yards of the opponent's endzone
- **`sec_in_half`**: The number of seconds remaining on the play clock in the current half
- **`epa_lag`**: Serial correlation of expected points added

\newpage

**_Team Characteristics:_**

- **`offense_formation`**: Categorical variable indicating what formation the offense uses
- **`receiver_alignment`**: Enumerated as 0x0, 1x0, 1x1, 2x0, 2x1, 2x2, 3x0, 3x1, 3x2 (text)
- **`curr_win_percentage`**: Win percentage for the current offensive team
- **`form_change`**: Uses shifts to determine if the offense changed formation before the ball was snapped


<!-- I'm not sure why the spacing for this last section isn't consistent with the first two -->
<!-- come back later to fix -->
**_Player Characteristics:_**

- **`wr_motion`**: Indicates if a wide receiver moved after the line was set
- **`te_motion`**: Indicates if a tight end moved after the line was set
- **`rb_motion`**: Indicates if a running back moved after the line was set
- **`wr_atsnap`**: Indicates if a wide receiver was in motion at the time of the snap
- **`te_atsnap`**: Indicates if a tight end was in motion at the time of the snap
- **`rb_atsnap`**: Indicates if a running back was in motion at the time of the snap

## 2. LASSO Model:

```{r, cache=TRUE}

#Load in the data set
nfl_df <- read.csv("data/modeling_data_frame.csv")

#Make the run pass variable a factor
nfl_df$run_pass <- as.factor(nfl_df$run_pass)

#Convert the boolean columns into numeric ones
bool_cols <- c("redzone", "wr_motion", "te_motion", "rb_motion",
               "wr_atsnap", "te_atsnap", "rb_atsnap")
nfl_df[bool_cols] <- lapply(nfl_df[bool_cols], as.numeric)

#Remove NA values
nfl_df <- nfl_df %>%
  drop_na()

#Use a lasso model to do varaible selection.

#Create a sequence of lambda values to try
lambda <- lambda <- 10^seq(-2, 10, length.out = 100)

#Create a recipe for the lasso model and normalize the predictors.
lasso_rec <- recipe(run_pass~., data=nfl_df) %>%
  
  #Remove unnecessary variables
  step_rm(game_id, play_id, possession_team,week) %>%
  
  step_novel(all_nominal_predictors()) %>%
  
  #Fix error for missing values by assigning them with unknown
  step_unknown(all_nominal_predictors()) %>%
  
  #Fix error where some variables are categorical and not numeric
  step_dummy(all_nominal_predictors()) %>%
  
  #Fix error from some of them having zero variance
  step_zv(all_predictors())%>%
  
  #Normalize the numeric predictors to have mean=0 and sd=1
  step_normalize(all_numeric_predictors())

#Create an empty data frame to store our lasso estimates
lasso_ests <- data.frame()

#Fit a lasso model for each lambda in our grid
for(lam in lambda){
  lasso_spec <- logistic_reg(mixture=1, penalty=lam)%>%
    set_mode("classification") %>%
    set_engine("glmnet")
  
  #Fit the model
  workflow() %>%
    add_model(lasso_spec) %>%
    add_recipe(lasso_rec) %>%
    fit(nfl_df) %>%
    tidy() %>%
    bind_rows(lasso_ests)-> lasso_ests
}

  #Create a line plot of 
lasso_ests |>
  filter(term != "(Intercept)") |>
  ggplot() +
  geom_line(aes(penalty, estimate, group = term, colour = term)) +
  coord_trans(x = "log10")+
  #The legend is so large it obstructs the graph. You can comment out this line
  # and view the graph in a new window to get a better idea. 
  theme(legend.position="none")
```

## 3. Tuning Lasso:

```{r}
#Fit a lasso model and determine the optimum lambda for varaible selection

#Create a data frame for tuning lambda
tune_df <- data.frame(lambda=lambda)

#Perform 10-fold cross-validation to get an estimate of test MSE for each
#labmda in our grid
nfl_10fld_cv <- vfold_cv(nfl_df, v=10)

#Specify the model
lasso_spec <- logistic_reg(mixture = 1, penalty = tune("lambda")) %>%
  set_mode("classification") %>%
  set_engine("glmnet")

#Create the workflow object for lasso
lasso_workflow <-workflow() %>%
  add_model(lasso_spec) %>%
  add_recipe(lasso_rec) 

#Perform the cross validation
tune_results <- tune_grid( lasso_workflow,
                           resamples=nfl_10fld_cv,
                           #For some reason it works if I let the labmda values
                           #be created automatically but not when I try to use
                           #the labmda values in the tuning 
                           grid=50)

#Determine the best lambda value
best_lambda <- select_best(tune_results, metric="accuracy") 

#Create the lasso model with the optimum lambda
final_lasso_workflow <- finalize_workflow(lasso_workflow,  best_lambda)

final_lasso_model <- fit(final_lasso_workflow, data=nfl_train) -> final_lasso_fit

final_model <- extract_fit_parsnip(final_lasso_model)

#Convert the coefficients to a tidy data frame
selected_variables <- tidy(final_model)

#Generate predictions on the test dataset
lasso_predictions <- predict(final_lasso_model, nfl_test, type = "class") %>%
  bind_cols(nfl_test)

#Create a confusion matrix
confusion_mat <- conf_mat(
  data = lasso_predictions,
  truth = run_pass, 
  estimate = .pred_class
)

#Print the confusion matrix
print(confusion_mat)
```

## 4. Logistic Regression Model (Entire Data Set):

```{r}

#Fit the logistic model on the entire data set using the varaibles from the 
#lasso model.

#Look at the output from the optimal lasso model to do variable selection
selected_vars <-c("down", "yards_to_go", "yards_to_endzone", "sec_in_half",
                  "epa_lag", "curr_win_percentage", "form_change", "wr_motion",
                  "te_motion","rb_motion","wr_atsnap", "te_atsnap",
                  "offense_formation", "receiver_alignment")

#Drop the few data points that are in the testing set but not the training set
nfl_train <- nfl_train %>% filter(offense_formation %in% nfl_test$offense_formation, 
                                      receiver_alignment %in% nfl_test$receiver_alignment)
  
nfl_test <- nfl_test %>% filter(offense_formation %in% nfl_train$offense_formation, 
                                    receiver_alignment %in% nfl_train$receiver_alignment)
  
#Create a formula for logistic regression using selected variables
logistic_formula <- as.formula(paste("run_pass ~",
                                     paste(selected_vars, collapse = " + ")))

#Fit the logistic regression model using the selected variables
final_logistic_model <- glm(
  formula = logistic_formula,
  data = nfl_train,
  family = binomial()
)

# Generate predictions on the test set
logistic_predictions_prob <- predict(final_logistic_model,
                                     newdata = nfl_test, type = "response")

#Convert probabilities to class labels based on a threshold of 0.5
logistic_predictions_class <- ifelse(logistic_predictions_prob > 0.5, "run", "pass")

#Add predictions to the test dataset
nfl_test <- nfl_test %>%
  mutate(predicted_run_pass = logistic_predictions_class)

#Create a confusion matrix
confusion_matrix <- table(
  Truth = nfl_test$run_pass,
  Prediction = nfl_test$predicted_run_pass
)

#Make sure that the precited pass column is also a factor like the truth col
nfl_test$predicted_run_pass <- as.factor(nfl_test$predicted_run_pass)

#Create a confusion matrix with the accuracy
confusionMatrix(nfl_test$predicted_run_pass, nfl_test$run_pass)
```

## 5. Logistic Regression Models by Team:

```{r, warning=FALSE}

#Fit the optimal logistic regression model for each team 


#Create a list to store the models for each team as well as the confusion
#matrices
model_info <- list("nfl_team" = unique(nfl_df$possession_team),
                   "train_fit" = as.list(seq(32)),
                   "prediction" = as.list(seq(32)),
                   "conf_mat" = as.list(seq(32)),
                   "m_corr"= as.list(seq(32)))

#Specify the logistic formula using the variables selected from lasso
logistic_formula <- as.formula(paste("run_pass ~",
                                   paste(selected_vars, collapse = " + ")))

#Use a for loop to create a model for each team and store it in the model info
#list
for (i in c(1:32)){
  
  #Grab the current team
  team<-model_info$nfl_team[i]
  
  #Add a print statement so we can keep track while its running
  print(paste("Working on team", team))
  
  #Create a data frame with the data for the current team
  team_df <- nfl_df %>%
    filter(possession_team==team) %>%
    mutate(run_pass = as.factor(run_pass))
  
  #troublesome receiver alignment value for denver remove it
  if(team=="DEN"){
    team_df <- team_df %>%
      filter(receiver_alignment !="2x0")
  }
  
  #Sort the data by week
  weeks <- sort(unique(team_df$week))
  
  #Split the data into training and testing
  team_train <- team_df %>%
    filter(week==c(1,2,3,4,5,6))
  
  team_test <- team_df %>%
    filter(week==c(7,8,9))
  
  #Drop the few data points that are in the testing set but not the training set
  team_train <- team_train %>% 
    filter(offense_formation %in% team_test$offense_formation, 
           receiver_alignment %in% team_test$receiver_alignment)

  team_test <- team_test %>% 
    filter(offense_formation %in% team_train$offense_formation,
           receiver_alignment %in% team_train$receiver_alignment)
  
  #Fit the logistic regression model using the selected variables
  final_logistic_model <- glm(
    formula = logistic_formula,
    data = nfl_train,
    family = binomial()
    )
  
  #Extract the model
  train_fit <- final_logistic_model$model
  
  #Add the training fit to the list
  model_info$train_fit[[i]] <- train_fit
  
  # Generate predictions on the test set
  logistic_predictions_prob <- predict(final_logistic_model,
                                       newdata = team_test, type = "response")
  
  #Convert probabilities to class labels based on a threshold of 0.5
  logistic_predictions_class <- ifelse(logistic_predictions_prob > 0.5,
                                       "run", "pass")
  
  #Add the predicted classes to the model info list
  model_info$predictions[[i]]<- logistic_predictions_class
  
  #Add predictions to the test dataset
  team_test <- team_test %>%
    mutate(predicted_run_pass = logistic_predictions_class)
  
  #Make sure that the precited pass column is also a factor like the truth col
  team_test$predicted_run_pass <- as.factor(team_test$predicted_run_pass)
  
  #Add the confusion matrix to the model info list
  model_info$conf_mat[[i]]<-confusionMatrix(team_test$predicted_run_pass,
                                            team_test$run_pass)
  
  #Calculate Matthews correlation coefficient (mltools)
  model_info$m_corr[[i]]<- mcc(preds=team_test$predicted_run_pass,
                               actuals= team_test$run_pass)
  
  #Add print statements to view the model progress
  
  #Print the first couple of confusion matrices as an example
  if(i<=2){
    print(confusionMatrix(team_test$predicted_run_pass,team_test$run_pass))
    print(model_info$m_corr[[i]])
  }
  
}
#Add an ending print statement
print("Complete")

#Save the data
saveRDS(model_info, "Logistic Model Info by Team.rds")
```
## 6. Visualizations:

```{r, warning=FALSE}

#Load in the data
log_model_info <- readRDS("Logistic Model Info by Team.rds")

#Convert some of the list objects into data frames for plotting
team_names <- as.data.frame(log_model_info$nfl_team)

#Rename the column
team_names <- team_names %>%
  rename(nfl_team = `log_model_info$nfl_team`)

mcc_vals <- as.data.frame(log_model_info$m_corr)

mcc_vals_long <- mcc_vals %>%
  pivot_longer(
    cols = everything(),        
    names_to = "variable",      
    values_to = "mcc_value"     
  )

#Combine the two into a single data frame for plotting
plot_df <- data.frame(Team = team_names$nfl_team,
                      MCC = mcc_vals_long$mcc_value)

#Use the correlations in the model info list to plot correllation by team
corr_plot <- plot_df %>%
  ggplot(aes(x=Team, y=MCC))+
  
  #Add the team logos for clarity
  geom_nfl_logos(aes(team_abbr = Team,
                     width = 0.075,
                     height=0.1))+
  
  #Add a horizontal line at y=0 to visualize the boundary
  geom_hline(yintercept = 0, color="red")+
  
  #Label the two points with the highest (best) MCC
  annotate("text", x="HOU", y=0.5, label="Best Agreement",
           color="steelblue", size=3)+
  annotate("segment", 
           x="DEN", xend="BUF",
           y=0.5, yend=0.53,
           size=2,colour="steelblue",
           alpha=0.6,arrow=arrow())+
  annotate("segment", 
           x="KC", xend="NE",
           y=0.5, yend=0.45,
           size=2,colour="steelblue",
           alpha=0.6,arrow=arrow())+
  
  #Label a point where MCC is approximately 0
  #i.e. teams where the model was as good as randomly guessing
  annotate("text", x="CIN", y=0.08, label="Random Guessing",
           color="steelblue", size= 3)+
  annotate("segment", x="DEN" ,xend="HOU", y=0.05, yend=0,
           size=2, alpha=0.6,color="steelblue",
           arrow=arrow())+
  
  #Label a point with negative agreement
  annotate("text", x="CIN", y=-0.08, label="Negative Agreement",
           color="steelblue", size=3)+
  annotate("segment", x="DET" ,xend="LV", y=-0.08, yend=-0.08,
           size=2, alpha=0.6,color="steelblue",
           arrow=arrow())+
  
  #Change the y-axis limits to be consistent with the other models MCC
  scale_y_continuous(limits=c(-0.3, 0.8))+
  
  #Add appropriate labels
  labs(x="Team",
       y="Matthew Correlation Coefficient (MCC)",
       title="Prediction on NFL Offenses- Run or Pass",
       subtitle="MCC of Logistic Regression (LASSO) ",
       caption="Data from: NFL Big Data Bowl (2025)")+
  
  theme_bw()+
  
  #Adjust plot elements for neatness
  theme(plot.title= element_text(hjust=0.5, size=20),
        plot.subtitle = element_text(hjust = 0.5, size=14),
        
        #Remove the x-axis labels and ticks to clean up the plot
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y= element_text(size=14),
        axis.title.x= element_text(size=14))

#Display the plot
corr_plot

#Save the plot and put it into the graphs folder
ggsave("graphs/MCC_Logistic.png", plot=corr_plot,
       width=12, height=7)

#Save the plot in a different size for the paper
ggsave("graphs/Images for paper/MCC_log.png", plot= corr_plot,
       height=4, width=6.5)
```
