---
title: "Reproduce Logistic Model (LASSO)"
author: "Justin C. Eldridge"
date: "2024-12-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Load the appropriate libraries:
library(ggplot2)
library(tidymodels)
library(tidyr)
library(glmnet)
library(mltools)
library(caret)
library(nflplotR)
library(dplyr)
library(tidyverse)
```

# Logistic Model (LASSO)

This document contains all the code necessary to reproduce the logistic models along with the associated visualizations.

## Contents:
  1. Variables
  2. Lasso Model
  3. Tuning Lasso
  4. Logistic Regression Model (Entire Data Set)
  5. Logistic Regression Models by Team
  6. Visualizations
  

## 1. Variables:

  Using our background knowledge and drawing on our EDA we decided to include the following variables in our models.
  
  
  **_Game Characteristics:_**

- **`game_id`**: Unique identifier for the game
- **`play_id`**: Unique identifier for a specific play
- **`possession_team`**: Identifies the offensive team
- **`week`**: Identifies which week of the season the game was in
- **`run_pass`**: Indicates if the play was a run or a pass (response variable)
- **`down`**: Identifies which down the offensive team was on for a particular play
- **`yards_to_go`**: The number of yards needed to get a first down
- **`yards_to_endzone`**: The number of yards to the opponent's endzone
- **`redzone`**: Indicates whether the offense is within 20 yards of the opponent's endzone
- **`sec_in_half`**: The number of seconds remaining on the play clock in the current half
- **`epa_lag`**: Serial correlation of expected points added

\newpage

**_Team Characteristics:_**

- **`offense_formation`**: Categorical variable indicating what formation the offense uses
- **`receiver_alignment`**: Enumerated as 0x0, 1x0, 1x1, 2x0, 2x1, 2x2, 3x0, 3x1, 3x2 (text)
- **`curr_win_percentage`**: Win percentage for the current offensive team
- **`form_change`**: Uses shifts to determine if the offense changed formation before the ball was snapped


<!-- I'm not sure why the spacing for this last section isn't consistent with the first two -->
<!-- come back later to fix -->
**_Player Characteristics:_**

- **`wr_motion`**: Indicates if a wide receiver moved after the line was set
- **`te_motion`**: Indicates if a tight end moved after the line was set
- **`rb_motion`**: Indicates if a running back moved after the line was set
- **`wr_atsnap`**: Indicates if a wide receiver was in motion at the time of the snap
- **`te_atsnap`**: Indicates if a tight end was in motion at the time of the snap
- **`rb_atsnap`**: Indicates if a running back was in motion at the time of the snap

## 2. LASSO Model:

```{r, cache=TRUE}

#Load in the data set
nfl_df <- read.csv("data/modeling_data_frame.csv")

#Make the run pass variable a factor
nfl_df$run_pass <- as.factor(nfl_df$run_pass)

#Convert the boolean columns into numeric ones
bool_cols <- c("redzone", "wr_motion", "te_motion", "rb_motion",
               "wr_atsnap", "te_atsnap", "rb_atsnap")
nfl_df[bool_cols] <- lapply(nfl_df[bool_cols], as.numeric)

#Remove NA values
nfl_df <- nfl_df %>%
  drop_na()

#Use a lasso model to do varaible selection.

#Create a sequence of lambda values to try
lambda <- lambda <- 10^seq(-2, 10, length.out = 100)

#Create a recipe for the lasso model and normalize the predictors.
lasso_rec <- recipe(run_pass~., data=nfl_df) %>%
  
  #Remove unnecessary variables
  step_rm(game_id, play_id, possession_team,week) %>%
  
  step_novel(all_nominal_predictors()) %>%
  
  #Fix error for missing values by assigning them with unknown
  step_unknown(all_nominal_predictors()) %>%
  
  #Fix error where some variables are categorical and not numeric
  step_dummy(all_nominal_predictors()) %>%
  
  #Fix error from some of them having zero variance
  step_zv(all_predictors())%>%
  
  #Normalize the numeric predictors to have mean=0 and sd=1
  step_normalize(all_numeric_predictors())

#Create an empty data frame to store our lasso estimates
lasso_ests <- data.frame()

#Fit a lasso model for each lambda in our grid
for(lam in lambda){
  lasso_spec <- logistic_reg(mixture=1, penalty=lam)%>%
    set_mode("classification") %>%
    set_engine("glmnet")
  
  #Fit the model
  workflow() %>%
    add_model(lasso_spec) %>%
    add_recipe(lasso_rec) %>%
    fit(nfl_df) %>%
    tidy() %>%
    bind_rows(lasso_ests)-> lasso_ests
}

  #Create a line plot of 
lasso_ests |>
  filter(term != "(Intercept)") |>
  ggplot() +
  geom_line(aes(penalty, estimate, group = term, colour = term)) +
  coord_trans(x = "log10")+
  #The legend is so large it obstructs the graph. You can comment out this line
  # and view the graph in a new window to get a better idea. 
  theme(legend.position="none")
```

